Testing the motor and cognitive foundations of Paleolithic social transmission

### 2021-08-16 manuscript/manuscript.Rmd
all figures and table added.

---
title: Testing the motor and cognitive foundations of Paleolithic social transmission
author: 
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
- Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret. Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability. However, we actually know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic. This paper presents results from a pilot experimental study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition. We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs. direct active teaching) employing appropriate raw materials, practice time, and real human interaction. Participant performance was evaluated through analysis of the stone artifacts produced. Performance was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study. Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group. These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.\\par \\textbf{Keywords:} Oldowan; Stone toolmaking; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

Sixty pounds of 3-to-5 inch basalt "Mexican Beach Pebbles" were purchased from a landscaping supply company for use as hammerstones in the experiment. Of these, 90 were selected as suitable for use. These weighed between 213g-1360g (mean = 425) and varied in elongation (L/W = 1.01 to 2.65) and relative thickness (LxW/T = 90.48 to 283.67). Forty-five stones were placed in the middle of the knapping area (Figure 2) for participants to freely choose from during the experiment. Broken hammerstones were replaced from the reserve to maintain a consistent number and range of choices. Each hammerstone was numbered and participants' choices were recorded along with the number of the core(s) being worked on with a particular hammerstone.

@@

All finished cores were weighed and measured (L, W, T). Delta weight was calculated as (Start weight-End weight)/Start weight. All detached pieces (DPs) were collected and weighed. We did not sort DPs into types (e.g., whole flakes, fragments) as this would have greatly increased processing time and it is not clear that such distinctions add relevant information regarding utility/desirability beyond that supplied by metrics [@stout2019]. All DPs larger than 40mm in maximum dimension were photographed and measured. It is conventional in Early Stone Age lithic analysis to employ a 20 mm cut-off. We selected a higher threshold for both pragmatic (analysis time) and theoretical reasons. Flake use experiments have shown that flakes weighing less than 5--10 g or with a surface area below 7--10 cm2 [@prasciunas2007] or with a maximum dimension \<50-60 mm [@key2014] become markedly inefficient for basic cutting tasks. Similarly, data from Oldowan replication experiments [@stout2019] show that the utility index (flake cutting edge/flake mass1/3) \* (1 - exp[-0.31 \* (flake maximum dimension -- 1.81)]) developed by Morgan et al. (2015) falls off rapidly below 40mm maximum dimension ( Mean Utility \< 40mm = 0.508; \>=40mm = 0.946; t= 11.99, df = 707, p \< 0.000). By including weight in our cut-off criteria we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.

@@

The analysis yielded three factors explaining a total of 90% of the variance for the entire 11 measurement variable set **(Table**). Factor 1 tracks flake size with higher scores indicating larger flakes since all 11 measures load positively on this factor. Factor 2's loadings track the increasing relationship between thickness, length, and flake width. As factor 2 scores increase, flakes get thicker, longer, and narrow, resembling irregular splinters. Factor 3 tracks the relationship between flake proximal and distal width relative to thickness. As factor 3 scores go up, flakes get thinner and narrower at the distal ends and wider at the base. Factor 3 therefore tracks flakes with a typical shape having a thin cross-section, wider base, and narrower tip. We used these three flake shape coordinates to approximate DP size and shape in the project's flake performance factor analysis.

@@

The analysis yielded two factors explaining a total of 56% of the variance for the entire set of variables (Table). Factor 1 (hereafter "Quantity") explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and on core delta mass. Performance factor 2 (hereafter "Quality") covers 27% of the sample variance and measures flaking quality as reflected in high positive loadings on Shape Factors 1 (size) and 3 (thin, "flake-like" shape) and a negative loading on Shape Factor 2 ("splinter-like" thickness and elongation). High scores on Quality thus reflect production of larger, relatively thinner, and more typically flake-shaped vs. splinter-shaped DPs.

@@

Thus, it appears that Trained participants achieved higher Total Productivity by increasing average flaking Quantity across cores and without sacrificing Quality, whereas Untrained participants found other ways to vary Total Productivity (e.g., number of cores knapped rather than Quantity per core, see variance Table and Figure) and generally increased productivity at the expense of Quality. Experimental artifacts illustrating these trade-offs are presented in Figure.

@@

Table summarizes the results of ANOVA tests group level difference in central tendency on various performance measures. We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors. In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (**Figure** \@ref(fig:fig7)). All three of these results show either medium or large effect sizes. In all three comparisons, the trained group's data distributions tended towards the expert sample although they were not significantly different from the untrained group (**Figure** \@ref(fig:fig8)). We also observed a significant difference in shape factor 2 (splinters) driven by the expert's lower values, but with a very low (\<0.01) effect size. These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.

@@

While we did not find significant differences in central tendency between our two experimental groups, results (**Figure** \@ref(fig:fig7)) did indicate lower variance in the trained group. To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the trained and untrained individuals using the F-test on either core-averaged or flake specific variances. Table and **Figure** \@ref(fig:fig9) present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass). In most instances, variance in the untrained group exceeds that of trained individuals by 1.5 to 4.7 times. The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the untrained group (**Figure** \@ref(fig:fig9)), rather than to shift the mean.

@@

Table shows no significant training effects across the two performance measures either as grouped data or between individuals (**Figure** \@ref(fig:fig10) and **Figure** \@ref(fig:fig11)). This result demonstrated that flaking outcomes did not change dramatically across the study interval. This lack of significant learning effects is confirmed by an inspection of individual learning curves (Figure). The one significant main training effect related to core starting mass (with a strong effect size = 0.25). On average, core starting masses start low and increase, showing that participants selected smaller nodules first. As the smaller nodules in their allotment were depleted, participants were left to knap larger, less preferred nodules. This preference for smaller cores is somewhat less pronounced in the untrained group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the untrained group (Figure???).

@@

*Flaking performance variable* $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength

@@

This model explains a statistically significant and substantial proportion of variance in flaking productivity ($R^2$ = 0.84, F (6, 14) = 12.7, p \< 0.01, adj. $R^2$ = 0.77). A model residuals normality test shows no significant differences with the normal distribution (p = 0.72) indicating that this relationship is linear. A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 2, df = 6, p = 0.8).

Table presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average. The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable. We found significant (p\< 0.05) and substantial (Standardized Estimate \>= 0.50, i.e. a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST. The main effect of Grip Strength (**Figure** \@ref(fig:fig12)), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.

```{r fig12, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Significant main effect of grip strength on overall flaking productivity."}
knitr::include_graphics(here::here("figure/Figure12.jpeg"))
```

@@

*Quantity* $\sim$ Highest n-back level + BEAST score + Fitt's score + Grip strength + Training condition$\times$Highest n-back level + Training condition$\times$BEAST score + Training condition$\times$Grip strength + Nodule mass (as control)

@@

@@

*Quality flaking* $\sim$ Highest n-back level + Fitt's score + Grip strength + Fitt's score$\times$BEAST score + Grip strength$\times$BEAST score + Grip strength$\times$Fitt's score + Training condition$\times$Grip strength + Nodule mass (as control)

This model explains a statistically significant and substantial proportion of variance in Quality ($R^2$ = 0.75, F (8, 12) = 4.6, p \< 0.01, adj. $R^2$ = 0.6) in the absence of any main training effects. A model residuals normality test shows no significant differences with the normal distribution (p = 0.41) indicating that this relationship is linear. A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 7, df = 8, p = 0.5).

@@

In our experiment, the strongest effects of teaching were to reduce variance (Figure, Table) rather than shift mean values. In particular, teaching acted as a "safety net" that homogenized performance by reducing the frequency of extremely poor outcomes (i.e., learning failures). This finding provides additional support for the hypothesis that teaching would have increased the reliability of Oldowan skill reproduction [@morgan2015] while simultaneously corroborating the view that basic flaking competence can be achieved in its absence [@tennie2017]. Our results thus do not imply that teaching was required or even present during Oldowan times (but see @gärdenfors2017), but rather serve to reinforce the plausibility of co-evolutionary scenarios positing the cost/reliability of technological skill acquisition as a selection pressure favoring the evolution of teaching and language [@stout2010; @morgan2015; @stout2017].

@@

# Figures

>>>>

---
title: Testing the motor and cognitive foundations of Paleolithic social transmission
author: 
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
- Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret. Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability. However, we actually know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic. This paper presents results from a pilot experimental study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition. We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs. direct active teaching) employing appropriate raw materials, practice time, and real human interaction. Participant performance was evaluated through analysis of the stone artifacts produced. Performance was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study. Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group. These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.\\par \\textbf{Keywords:} Oldowan; Stone toolmaking; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

Sixty pounds of 3-to-5 inch basalt "Mexican Beach Pebbles" were purchased from a landscaping supply company for use as hammerstones in the experiment. Of these, 90 were selected as suitable for use. These weighed between 213g-1360g (mean = 425) and varied in elongation (L/W = 1.01 to 2.65) and relative thickness (L$\times$W/T = 90.48 to 283.67). Forty-five stones were placed in the middle of the knapping area (**Figure** \@ref(fig:fig3)) for participants to freely choose from during the experiment. Broken hammerstones were replaced from the reserve to maintain a consistent number and range of choices. Each hammerstone was numbered and participants' choices were recorded along with the number of the core(s) being worked on with a particular hammerstone.

@@

**Table** \@ref(tab:tab1) lists details for the study’s various lithic attributes. All finished cores were weighed and measured (L, W, T). Delta weight was calculated as (Start weight-End weight)/Start weight. All detached pieces (DPs) were collected and weighed. We did not sort DPs into types (e.g., whole flakes, fragments) as this would have greatly increased processing time and it is not clear that such distinctions add relevant information regarding utility/desirability beyond that supplied by metrics [@stout2019]. All DPs larger than 40mm in maximum dimension were photographed and measured. It is conventional in Early Stone Age lithic analysis to employ a 20 mm cut-off. We selected a higher threshold for both pragmatic (analysis time) and theoretical reasons. Flake use experiments have shown that flakes weighing less than 5--10 g or with a surface area below 7--10 cm2 [@prasciunas2007] or with a maximum dimension \<50-60 mm [@key2014] become markedly inefficient for basic cutting tasks. Similarly, data from Oldowan replication experiments [@stout2019] show that the utility index (flake cutting edge/flake mass1/3) \* (1 - exp[-0.31 \* (flake maximum dimension -- 1.81)]) developed by Morgan et al. (2015) falls off rapidly below 40mm maximum dimension ( Mean Utility \< 40mm = 0.508; \>=40mm = 0.946; t= 11.99, df = 707, p \< 0.000). By including weight in our cut-off criteria we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.
```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1 <- data.frame("Variable" = c("Delta mass", "Mass of large detached pieces / total debitage mass", "Length", "Width", "Thickness"), 
                 "Variable class" = c("Ratio", "Ratio", "Linear measurement", "Linear measurement", "Linear measurement"), 
                 "Definition" = c("Final core mass as a percentage of original nodule mass. Higher values show more completely flaked nodules.", "Combined mass of all detached pieces >40mm in maximum dimension and 5g in mass divided by the mass of all material detached from a specific core. Higher values show cores with more detached pieces per knapped material.", "Measurement along each core and flake's longest axis in mm.", "Measurement taken at 10% increments along a detached piece's longest axis starting at the detached piece's tip (in mm). Same measurement taken orthogonal to a core's length.", "Measurement taken at the maximum point in a detached piece's profile and orthogonal to width on the cores."),
                "Recording method" = c("Digital scale", "Calipers, counts, digital scale", "2D photogrammetry", "2D photogrammetry", "Calipers") )

kableExtra::kbl(df1, booktabs = T, caption = "Overview of lithic variables and recording methods used in the study.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7) %>%
  kableExtra::column_spec(1, width = "10em") %>%
  kableExtra::column_spec(3, width = "20em")%>%
  kableExtra::column_spec(4, width = "10em")

```

@@

The analysis yielded three factors explaining a total of 90% of the variance for the entire 11 measurement variable set (**Table** \@ref(tab:tab2)). Factor 1 tracks flake size with higher scores indicating larger flakes since all 11 measures load positively on this factor. Factor 2's loadings track the increasing relationship between thickness, length, and flake width. As factor 2 scores increase, flakes get thicker, longer, and narrow, resembling irregular splinters. Factor 3 tracks the relationship between flake proximal and distal width relative to thickness. As factor 3 scores go up, flakes get thinner and narrower at the distal ends and wider at the base. Factor 3 therefore tracks flakes with a typical shape having a thin cross-section, wider base, and narrower tip. We used these three flake shape coordinates to approximate DP size and shape in the project's flake performance factor analysis.
```{r tab2, echo=FALSE, message=FALSE}
library(tidyverse)
df2<-read.csv(here::here("data/tables/table2.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df2, booktabs = T, caption = "Lithic size/shape PCA factor loadings.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

@@

The analysis yielded two factors explaining a total of 56% of the variance for the entire set of variables (**Table** \@ref(tab:tab3)). Factor 1 (hereafter "Quantity") explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and on core delta mass. Performance factor 2 (hereafter "Quality") covers 27% of the sample variance and measures flaking quality as reflected in high positive loadings on Shape Factors 1 (size) and 3 (thin, "flake-like" shape) and a negative loading on Shape Factor 2 ("splinter-like" thickness and elongation). High scores on Quality thus reflect production of larger, relatively thinner, and more typically flake-shaped vs. splinter-shaped DPs.
```{r tab3, echo=FALSE, message=FALSE}
library(tidyverse)
df3<-read.csv(here::here("data/tables/table3.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df3, booktabs = T, caption = "Overview of principal components results on the six flake performance measures. DP = detached piece.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

@@

Thus, it appears that Trained participants achieved higher Total Productivity by increasing average flaking Quantity across cores and without sacrificing Quality, whereas Untrained participants found other ways to vary Total Productivity (e.g., number of cores knapped rather than Quantity per core, see variance Table and Figure) and generally increased productivity at the expense of Quality. Experimental artifacts illustrating these trade-offs are presented in **Figure???**.

@@

**Table** \@ref(tab:tab4) summarizes the results of ANOVA tests group level difference in central tendency on various performance measures. We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors. In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (**Figure** \@ref(fig:fig7)). All three of these results show either medium or large effect sizes. In all three comparisons, the trained group's data distributions tended towards the expert sample although they were not significantly different from the untrained group (**Figure** \@ref(fig:fig8)). We also observed a significant difference in shape factor 2 (splinters) driven by the expert's lower values, but with a very low (\<0.01) effect size. These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.
```{r tab4, echo=FALSE, message=FALSE}
library(tidyverse)
df4<-read.csv(here::here("data/tables/table4.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df4, booktabs = T, caption = "Summary group-wise comparison statistics contrasting flake performance metrics by experiment condition and with the expert instructor. All comparisons are three-way except for the two skill PC factors and the total mass removed, which are two-way comparisons between the subject population as these factors were not applied to the expert data sample. Eta2 = ANOVA effect size (>0.1 = medium effect size, >0.2 = large effect size). * = non-parametric permuted p-values to account for unequal sample variances. SS = sum of squared differences, df = degrees of freedom, MS = mean squared difference.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

@@

While we did not find significant differences in central tendency between our two experimental groups, results (**Figure** \@ref(fig:fig7)) did indicate lower variance in the trained group. To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the trained and untrained individuals using the F-test on either core-averaged or flake specific variances. **Table** \@ref(tab:tab5) and **Figure** \@ref(fig:fig9) present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass). In most instances, variance in the untrained group exceeds that of trained individuals by 1.5 to 4.7 times. The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the untrained group (**Figure** \@ref(fig:fig9)), rather than to shift the mean.
```{r tab5, echo=FALSE, message=FALSE}
library(tidyverse)
df5<-read.csv(here::here("data/tables/table5.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df5, booktabs = T, caption = "F-test results comparing variances between trained and untrained subjects across 
our various flaking performance metrics.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

@@

**Table** \@ref(tab:tab6) shows no significant training effects across the two performance measures either as grouped data or between individuals (**Figure** \@ref(fig:fig10) and **Figure** \@ref(fig:fig11)). This result demonstrated that flaking outcomes did not change dramatically across the study interval. This lack of significant learning effects is confirmed by an inspection of individual learning curves (Figure). The one significant main training effect related to core starting mass (with a strong effect size = 0.25). On average, core starting masses start low and increase, showing that participants selected smaller nodules first. As the smaller nodules in their allotment were depleted, participants were left to knap larger, less preferred nodules. This preference for smaller cores is somewhat less pronounced in the untrained group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the untrained group (Figure???).
```{r tab6, echo=FALSE, message=FALSE}
library(tidyverse)
df6<-read.csv(here::here("data/tables/table6.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df6, booktabs = T, caption = "Summary group-wise comparison statistics contrasting flake performance metrics by experiment condition and relative core order. All comparisons are three-way. Eta2 = ANOVA effect size (>0.1 = medium effect size, >0.2 = large effect size). SS = sum of squared differences, df = degrees of freedom, MS = mean squared difference.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

@@

*Flaking performance variable $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength*

@@

This model explains a statistically significant and substantial proportion of variance in flaking productivity ($R^2$ = 0.84, F (6, 14) = 12.7, p \< 0.01, adj. $R^2$ = 0.77). A model residuals normality test shows no significant differences with the normal distribution (p = 0.72) indicating that this relationship is linear. A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 2, df = 6, p = 0.8).
```{r tab7, echo=FALSE, message=FALSE}
library(tidyverse)
df7<-read.csv(here::here("data/tables/table7.csv"), fileEncoding = 'UTF-8-BOM')
colnames(df7)[colnames(df7) %in% c("X95..CI", "t..12.")] <- c("95% CI", "t(12)")
kableExtra::kbl(df7, booktabs = T, caption = "Summary and estimates for the overall flaking productivity model.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

**Table** \@ref(tab:tab7) presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average. The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable. We found significant (p\< 0.05) and substantial (Standardized Estimate \>= 0.50, i.e. a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST. The main effect of Grip Strength (**Figure** \@ref(fig:fig12)), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.
```{r fig12, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Significant main effect of grip strength on overall flaking productivity."}
knitr::include_graphics(here::here("figure/Figure12.jpeg"))
```

@@

*Quantity $\sim$ Highest n-back level + BEAST score + Fitt's score + Grip strength + Training condition$\times$Highest n-back level + Training condition$\times$BEAST score + Training condition$\times$Grip strength + Nodule mass (as control)*

@@

**Table** \@ref(tab:tab8) presents this model’s coefficients and summary outputs, following the same format as Table. 
```{r tab8, echo=FALSE, message=FALSE}
library(tidyverse)
df8<-read.csv(here::here("data/tables/table8.csv"), fileEncoding = 'UTF-8-BOM')
colnames(df8)[colnames(df8) %in% c("X95..CI", "t..12.")] <- c("95% CI", "t(12)")
kableExtra::kbl(df8, booktabs = T, caption = "Summary and estimates for Quantity flaking model.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

@@

*Quality flaking $\sim$ Highest n-back level + Fitt's score + Grip strength + Fitt's score$\times$BEAST score + Grip strength$\times$BEAST score + Grip strength$\times$Fitt's score + Training condition$\times$Grip strength + Nodule mass (as control)*

This model explains a statistically significant and substantial proportion of variance in Quality ($R^2$ = 0.75, F (8, 12) = 4.6, p \< 0.01, adj. $R^2$ = 0.6) in the absence of any main training effects. A model residuals normality test shows no significant differences with the normal distribution (p = 0.41) indicating that this relationship is linear. A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 7, df = 8, p = 0.5).
```{r tab9, echo=FALSE, message=FALSE}
library(tidyverse)
df9<-read.csv(here::here("data/tables/table9.csv"), fileEncoding = 'UTF-8-BOM')
colnames(df9)[colnames(df9) %in% c("X95..CI", "t..12.")] <- c("95% CI", "t(12)")
kableExtra::kbl(df9, booktabs = T, caption = "Summary and estimates for the Quality flaking model.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

**Table** \@ref(tab:tab9) presents this model’s coefficients and summary outputs following the same data format as **Table** \@ref(tab:tab7) & **Table** \@ref(tab:tab8). The results show no effects that were both significant and substantial (Estimate >= 0.5).

@@

In our experiment, the strongest effects of teaching were to reduce variance (**Figure** \@ref(fig:fig9), **Table** \@ref(tab:tab5)) rather than shift mean values. In particular, teaching acted as a "safety net" that homogenized performance by reducing the frequency of extremely poor outcomes (i.e., learning failures). This finding provides additional support for the hypothesis that teaching would have increased the reliability of Oldowan skill reproduction [@morgan2015] while simultaneously corroborating the view that basic flaking competence can be achieved in its absence [@tennie2017]. Our results thus do not imply that teaching was required or even present during Oldowan times (but see @gärdenfors2017), but rather serve to reinforce the plausibility of co-evolutionary scenarios positing the cost/reliability of technological skill acquisition as a selection pressure favoring the evolution of teaching and language [@stout2010; @morgan2015; @stout2017].

@@
